---
title: "BLAST Primer Specificity Report"
subtitle: "Sea Otter SNP Genotyping Panel"
author: "Ellen Dymit"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Settings

Edit these before knitting if your file paths or thresholds differ.

```{r settings, echo=TRUE}
INPUT_CSV      <- "summary.csv"   # path to your BLAST summary CSV
OUTDIR         <- "blast_figures" # output folder for saved CSVs
PCT_CUTOFF_LO  <- 85              # conservative threshold (3 mismatches in ~20 bp)
PCT_CUTOFF_HI  <- 90              # stringent threshold (2 mismatches in ~20 bp)
PCT_CUTOFF_PERFECT <- 100         # perfect match - highest concern
EVALUE_CUTOFF  <- 0.01            # e-value threshold (secondary filter)
TARGET_SPECIES <- "sea_otter"     # should have hits - that's expected
```

---

## Setup

```{r libraries}
library(tidyverse)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

dir.create(OUTDIR, showWarnings = FALSE)
```

```{r load-data}
raw <- read_csv(INPUT_CSV, show_col_types = FALSE)

# Separate real hits from no-hit placeholders
hits <- raw %>%
  filter(hit_id != "NO_HIT") %>%
  mutate(
    pct_identity   = as.numeric(pct_identity),
    hsp_evalue     = as.numeric(hsp_evalue),
    spans_junction = as.logical(spans_junction),
    is_nontarget   = species != TARGET_SPECIES
  )

all_pairs         <- sort(unique(raw$pair_name))
all_species       <- sort(unique(raw$species))
nontarget_species <- setdiff(all_species, TARGET_SPECIES)
```

Loaded **`r nrow(raw)`** rows covering **`r n_distinct(raw$pair_name)`** primer pairs 
across **`r n_distinct(raw$species)`** species 
(`r paste(sort(unique(raw$species)), collapse = ", ")`).

---

## Summary Table

Three identity flags are computed for each pair:

- **flag_85** - any non-target hit with pct_identity > `r PCT_CUTOFF_LO`% (conservative; 3 mismatches in a 20 bp primer)
- **flag_90** - any non-target hit with pct_identity > `r PCT_CUTOFF_HI`% (stringent; 2 mismatches in a 20 bp primer)
- **flag_100** - any non-target hit with pct_identity = 100% (perfect match; highest concern)

The `recommendation` column uses 100% as "CRITICAL", 90% as "MANUAL CHECK", 85-90% as "REVIEW".

```{r pair-summary}
pair_summary <- expand_grid(pair_name = all_pairs) %>%
  left_join(
    raw %>%
      group_by(pair_name) %>%
      summarise(forward = first(forward), reverse = first(reverse), .groups = "drop"),
    by = "pair_name"
  ) %>%
  left_join(
    hits %>%
      filter(is_nontarget) %>%
      group_by(pair_name) %>%
      summarise(
        worst_nontarget_pct    = max(pct_identity, na.rm = TRUE),
        worst_nontarget_evalue = min(hsp_evalue,   na.rm = TRUE),
        any_junction_nontarget = any(spans_junction, na.rm = TRUE),
        nontarget_species_hit  = paste(sort(unique(species)), collapse = ", "),
        .groups = "drop"
      ),
    by = "pair_name"
  ) %>%
  left_join(
    hits %>%
      filter(species == TARGET_SPECIES) %>%
      group_by(pair_name) %>%
      summarise(
        target_best_pct    = max(pct_identity, na.rm = TRUE),
        target_best_evalue = min(hsp_evalue,   na.rm = TRUE),
        .groups = "drop"
      ),
    by = "pair_name"
  ) %>%
  replace_na(list(
    worst_nontarget_pct    = 0,
    worst_nontarget_evalue = 99,
    any_junction_nontarget = FALSE,
    nontarget_species_hit  = "none",
    target_best_pct        = 0,
    target_best_evalue     = 99
  )) %>%
  mutate(
    no_target_hit  = target_best_pct == 0,
    flag_85        = worst_nontarget_pct > PCT_CUTOFF_LO | any_junction_nontarget,
    flag_90        = worst_nontarget_pct > PCT_CUTOFF_HI | any_junction_nontarget,
    flag_100       = worst_nontarget_pct >= PCT_CUTOFF_PERFECT,
    recommendation = case_when(
      no_target_hit & (any_junction_nontarget | worst_nontarget_pct > PCT_CUTOFF_HI) ~
        "CAUTION - no target hit AND concerning non-target hits",
      no_target_hit ~
        "CAUTION - no hit against target species (sea otter)",
      flag_100 & any_junction_nontarget ~
        "CRITICAL - perfect match + junction span in non-target",
      flag_100 ~
        "CRITICAL - perfect match (100% identity) in non-target",
      any_junction_nontarget & worst_nontarget_pct > PCT_CUTOFF_HI ~
        "MANUAL CHECK - >90% identity + junction span in non-target",
      any_junction_nontarget ~
        "MANUAL CHECK - junction span detected in non-target",
      worst_nontarget_pct > PCT_CUTOFF_HI ~
        "MANUAL CHECK - pct_identity > 90% in non-target",
      worst_nontarget_pct > PCT_CUTOFF_LO ~
        "REVIEW - pct_identity 85-90% in non-target",
      worst_nontarget_evalue < EVALUE_CUTOFF ~
        "REVIEW - significant e-value in non-target (but low identity)",
      TRUE ~ "PROCEED - no concerning non-target hits"
    )
  ) %>%
  arrange(desc(flag_100), desc(flag_90), desc(flag_85), desc(worst_nontarget_pct))

flagged_100     <- pair_summary %>% filter(flag_100)
flagged_90_only <- pair_summary %>% filter(flag_90 & !flag_100 & !no_target_hit)
flagged_85_only <- pair_summary %>% filter(flag_85 & !flag_90 & !no_target_hit)
clean           <- pair_summary %>% filter(!flag_85 & !no_target_hit)

write_csv(pair_summary,    file.path(OUTDIR, "all_pairs_summary.csv"))
write_csv(flagged_100,     file.path(OUTDIR, "critical_pairs_100pct.csv"))
write_csv(flagged_90_only, file.path(OUTDIR, "flagged_pairs_90pct.csv"))
write_csv(flagged_85_only, file.path(OUTDIR, "review_pairs_85to90pct.csv"))
write_csv(clean,           file.path(OUTDIR, "clean_pairs.csv"))
```

Quick counts:

| Category | n |
|---|---|
| Total primer pairs | `r nrow(pair_summary)` |
| No hit against sea otter | `r sum(pair_summary$no_target_hit)` |
| **CRITICAL: perfect match (100%)** in non-target | `r nrow(flagged_100)` |
| **Flagged at 90%** (manual Primer-BLAST check recommended) | `r nrow(flagged_90_only)` |
| **Flagged at 85% only** (borderline; worth a look) | `r nrow(flagged_85_only)` |
| Proceed (no hits above 85% in non-target) | `r nrow(clean)` |

The full summary table is below and saved to `blast_figures/all_pairs_summary.csv`.

```{r summary-table}
rec_colors <- c(
  "CAUTION - no target hit AND concerning non-target hits"    = "#cc0000",
  "CAUTION - no hit against target species (sea otter)"       = "#e06666",
  "CRITICAL - perfect match + junction span in non-target"    = "#4a0000",
  "CRITICAL - perfect match (100% identity) in non-target"   = "#7b0000",
  "MANUAL CHECK - >90% identity + junction span in non-target" = "#f6a623",
  "MANUAL CHECK - junction span detected in non-target"       = "#f6a623",
  "MANUAL CHECK - pct_identity > 90% in non-target"          = "#f6a623",
  "REVIEW - pct_identity 85-90% in non-target"               = "#ffd966",
  "REVIEW - significant e-value in non-target (but low identity)" = "#d9ead3",
  "PROCEED - no concerning non-target hits"                   = "#b7e1cd"
)

pair_summary %>%
  select(pair_name, no_target_hit, flag_85, flag_90, flag_100,
         worst_nontarget_pct, any_junction_nontarget,
         nontarget_species_hit, recommendation) %>%
  rename(
    `Pair`                 = pair_name,
    `No target hit`        = no_target_hit,
    `Flag >85%`            = flag_85,
    `Flag >90%`            = flag_90,
    `Flag 100%`            = flag_100,
    `Worst non-target %id` = worst_nontarget_pct,
    `Junction hit`         = any_junction_nontarget,
    `Non-target species`   = nontarget_species_hit,
    `Recommendation`       = recommendation
  ) %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12) %>%
  column_spec(
    9,
    background = rec_colors[pair_summary$recommendation],
    color = ifelse(pair_summary$recommendation %in%
                     c("CAUTION - no target hit AND concerning non-target hits",
                       "CAUTION - no hit against target species (sea otter)",
                       "CRITICAL - perfect match + junction span in non-target",
                       "CRITICAL - perfect match (100% identity) in non-target"),
                   "white", "black")
  )
```

---

## Figure 1: Max % Identity Heatmap

Max BLAST % identity per primer pair x species. Numbers in each cell show the 
highest % identity hit found. The color scale transitions from green through yellow 
at `r PCT_CUTOFF_LO`% and red at `r PCT_CUTOFF_HI`%. The sea otter column (target) 
is outlined in blue. Pairs are sorted by worst non-target hit (cleanest at top).

```{r heatmap-identity, fig.width=8, fig.height=max(5, length(all_pairs) * 0.22 + 2)}
max_identity <- hits %>%
  group_by(pair_name, species) %>%
  summarise(max_pct = max(pct_identity, na.rm = TRUE), .groups = "drop")

grid <- expand_grid(pair_name = all_pairs, species = all_species) %>%
  left_join(max_identity, by = c("pair_name", "species")) %>%
  replace_na(list(max_pct = 0))

pair_order <- grid %>%
  filter(species != TARGET_SPECIES) %>%
  group_by(pair_name) %>%
  summarise(worst_nontarget = max(max_pct)) %>%
  arrange(worst_nontarget, pair_name) %>%
  pull(pair_name)

grid <- grid %>%
  mutate(pair_name = factor(pair_name, levels = pair_order))

ggplot(grid, aes(x = species, y = pair_name, fill = max_pct)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(
    data = filter(grid, max_pct > 0),
    aes(label = round(max_pct, 0)),
    size = 2.5, color = "white", fontface = "bold"
  ) +
  scale_fill_gradientn(
    colours = c("#d9ead3", "#ffd966", "#e06666", "#cc0000", "#4a0000"),
    values  = rescale(c(0, PCT_CUTOFF_LO, PCT_CUTOFF_HI, 99, 100)),
    limits  = c(0, 100),
    name    = "Max % identity\n(best BLAST hit)"
  ) +
  geom_vline(
    xintercept = which(sort(all_species) == TARGET_SPECIES) - 0.5,
    color = "grey40", linetype = "dashed", linewidth = 0.6
  ) +
  annotate("rect",
    xmin = which(sort(all_species) == TARGET_SPECIES) - 0.5,
    xmax = which(sort(all_species) == TARGET_SPECIES) + 0.5,
    ymin = 0.5, ymax = length(all_pairs) + 0.5,
    fill = NA, color = "steelblue", linewidth = 0.8
  ) +
  labs(
    title    = "Max BLAST % Identity per Primer Pair x Species",
    subtitle = paste0(
      "Blue outline = target species (", TARGET_SPECIES, ").  ",
      "Yellow = >", PCT_CUTOFF_LO, "%; red = >", PCT_CUTOFF_HI, "%; dark red = 100%.\n",
      "Pairs sorted by worst non-target hit (cleanest at top)."
    ),
    x = "Species", y = "Primer Pair"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y      = element_text(size = 7),
    axis.text.x      = element_text(angle = 30, hjust = 1),
    plot.title       = element_text(face = "bold"),
    plot.subtitle    = element_text(size = 8, color = "grey40"),
    legend.position  = "right",
    panel.grid       = element_blank()
  )
```

---
---

## Critical Pairs (100% identity in non-target)

These **`r nrow(flagged_100)`** pairs had at least one perfect-match BLAST hit in a
non-target species. These are the highest priority to investigate or drop.
Also saved to `blast_figures/critical_pairs_100pct.csv`.

Note that this does not guarantee that non-target PCR amplification will occur.
Forward and reverse binding sites may be too far apart for an amplicon to form.
This script also ignores strand orientation and whether both primers match (vs. one).

```{r critical-table}
if (nrow(flagged_100) > 0) {
  flagged_100 %>%
    select(pair_name, forward, reverse, no_target_hit, flag_85, flag_90, flag_100,
           worst_nontarget_pct, any_junction_nontarget,
           nontarget_species_hit, recommendation) %>%
    rename(
      `Pair`                 = pair_name,
      `Forward`              = forward,
      `Reverse`              = reverse,
      `No target hit`        = no_target_hit,
      `Flag >85%`            = flag_85,
      `Flag >90%`            = flag_90,
      `Flag 100%`            = flag_100,
      `Worst non-target %id` = worst_nontarget_pct,
      `Junction hit`         = any_junction_nontarget,
      `Non-target species`   = nontarget_species_hit,
      `Recommendation`       = recommendation
    ) %>%
    kable(digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 12)
} else {
  cat("No pairs had a perfect-match hit in a non-target species.")
}
```

---

## Manual Check Pairs (90-99% identity)

These **`r nrow(flagged_90_only)`** pairs had non-target hits above 90% but below
100% identity (or junction hits), and no perfect-match hit. Run through
[Primer-BLAST](https://www.ncbi.nlm.nih.gov/tools/primer-blast/) to confirm
whether they would actually form an amplicon.
Also saved to `blast_figures/flagged_pairs_90pct.csv`.

```{r flagged-table}
if (nrow(flagged_90_only) > 0) {
  flagged_90_only %>%
    select(pair_name, forward, reverse, no_target_hit, flag_85, flag_90, flag_100,
           worst_nontarget_pct, any_junction_nontarget,
           nontarget_species_hit, recommendation) %>%
    rename(
      `Pair`                 = pair_name,
      `Forward`              = forward,
      `Reverse`              = reverse,
      `No target hit`        = no_target_hit,
      `Flag >85%`            = flag_85,
      `Flag >90%`            = flag_90,
      `Flag 100%`            = flag_100,
      `Worst non-target %id` = worst_nontarget_pct,
      `Junction hit`         = any_junction_nontarget,
      `Non-target species`   = nontarget_species_hit,
      `Recommendation`       = recommendation
    ) %>%
    kable(digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 12)
} else {
  cat("No pairs fall in the 90-99% range.")
}
```

---

## Review Pairs (85-90% only)

These **`r nrow(flagged_85_only)`** pairs had non-target hits between 85-90% identity
but nothing above 90% and no junction hits. They are lower priority but worth a
second look if your final panel needs to be especially stringent.
Also saved to `blast_figures/review_pairs_85to90pct.csv`.

```{r review-table}
if (nrow(flagged_85_only) > 0) {
  flagged_85_only %>%
    select(pair_name, forward, reverse, worst_nontarget_pct,
           nontarget_species_hit, recommendation) %>%
    rename(
      `Pair`                 = pair_name,
      `Forward`              = forward,
      `Reverse`              = reverse,
      `Worst non-target %id` = worst_nontarget_pct,
      `Non-target species`   = nontarget_species_hit,
      `Recommendation`       = recommendation
    ) %>%
    kable(digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 12)
} else {
  cat("No pairs fall in the 85-90% borderline range.")
}
```

---

## Clean Pairs

These **`r nrow(clean)`** pairs had no non-target hits above 85% and at least one
sea otter BLAST hit. Also saved to `blast_figures/clean_pairs.csv`.

```{r clean-table}
clean %>%
  select(pair_name, forward, reverse, target_best_pct,
         worst_nontarget_pct, recommendation) %>%
  rename(
    `Pair`                 = pair_name,
    `Forward`              = forward,
    `Reverse`              = reverse,
    `Best sea otter %id`   = target_best_pct,
    `Worst non-target %id` = worst_nontarget_pct,
    `Recommendation`       = recommendation
  ) %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12)
```

---

*Generated with `analyze_blast_results.Rmd` on `r Sys.Date()`.*
